---
title: "Born under crisis: Early-life conditions and health outcomes during the Finnish War"
author:
 - name: "Erling Häggström Gunfridsson"
   affiliation: "Umeå University"
 - name: "Elina Nordun Omnell"
   affiliation: "Umeå University"
date: today
format:
 html:
   toc: true
   number-sections: true
 pdf:
   toc: true
   number-sections: true
   documentclass: article
   geometry: margin=2.5cm
   fontsize: 11pt
execute:
 echo: false
 warning: false
 message: false
 dir: project
---

## Abstract {-}

This paper examines the long-term consequences of early-life exposure to crisis conditions during the Finnish War (1808–1809). Using longitudinal individual-level data from northern Sweden, we analyse both survival and early-life health outcomes. The results indicate that cohorts born during the crisis exhibit distinct patterns in mortality and disability, consistent with both selection effects and causal impacts of adverse early-life environments.

**Keywords:** early-life conditions, mortality, health, Finnish War, historical demography


## Short bio {-}
Erling Häggström Gunfridsson and Elina Nordung Omnell are researchers at the Centre for
Demographic and Ageing Research (CEDAR) and the Demographic Data Base (DDB) at Umeå
University. Their work centers on historical demography, with a particular focus on mortality, and
reproductive patterns in pre-industrial Sweden.
A core mission of CEDAR and DDB is the digitization and systematic linkage of historical Swedish
parish records. Through record linkage, individuals and families can be followed longitudinally
across time and place, enabling detailed analyses of demographic behavior, social stratification, and
long-term health outcomes. Their research combines this large-scale longitudinal data to examine
how crises and social inequality shape life courses and population dynamics.

```{r}
library(targets)
library(data.table)
library(ggplot2)
library(knitr)

case_name <- "war"
store_path <- "_targets"

pick_case_table <- function(x, case_name = "war") {
  dt <- data.table::as.data.table(x)
  if (!"case" %in% names(dt)) return(dt)
  out <- dt[case == case_name]
  if (nrow(out) == 0L) {
    stop(sprintf("No rows found for case '%s'.", case_name), call. = FALSE)
  }
  out
}

format_p_value <- function(p) {
  ifelse(
    is.na(p),
    NA_character_,
    ifelse(p < 0.001, "<0.001", sprintf("%.3f", p))
  )
}

format_lyte_table <- function(dt, digits = 3) {
  dt <- data.table::copy(data.table::as.data.table(dt))

  keep <- intersect(
    c(
      "case", "term", "estimate", "std.error", "statistic", "p.value",
      "odds_ratio", "conf.low.or", "conf.high.or"
    ),
    names(dt)
  )
  dt <- dt[, ..keep]

  if ("term" %in% names(dt)) {
    dt <- dt[term != "(Intercept)"]
  }

  num_cols <- intersect(
    c("estimate", "std.error", "statistic", "odds_ratio", "conf.low.or", "conf.high.or"),
    names(dt)
  )

  if (length(num_cols) > 0) {
    dt[, (num_cols) := lapply(.SD, round, digits), .SDcols = num_cols]
  }

  if ("p.value" %in% names(dt)) {
    dt[, p.value := format_p_value(p.value)]
  }

  dt
}

format_global_tests <- function(dt, digits = 2) {
  dt <- data.table::copy(data.table::as.data.table(dt))

  keep <- intersect(c("case", "term", "df", "statistic", "p.value"), names(dt))
  dt <- dt[, ..keep]

  num_cols <- intersect(c("df", "statistic"), names(dt))

  if (length(num_cols) > 0) {
    dt[, (num_cols) := lapply(.SD, round, digits), .SDcols = num_cols]
  }

  if ("p.value" %in% names(dt)) {
    dt[, p.value := format_p_value(p.value)]
  }

  dt
}

clean_term_labels <- function(x) {
  x <- gsub("^factor\\(fod_ar\\)", "Birth year: ", x)
  x <- gsub("^fod_ar$", "Birth year", x)
  x <- gsub("^oaktaTRUE$", "Illegitimate birth", x)
  x <- gsub("^oakta$", "Illegitimate birth", x)
  x <- gsub("^kon$", "Sex", x)
  x <- gsub("^kon1$", "Sex: category 1", x)
  x <- gsub("^kon2$", "Sex: category 2", x)
  x
}

image_md <- function(path) {
  uri <- knitr::image_uri(path)
  knitr::asis_output(sprintf("![](%s)", uri))
}

# tables
desc_all <- tar_read(descriptive_table, store = store_path)
desc_war <- pick_case_table(desc_all, case_name)

lyte_tbl_all <- tar_read(lyte_model_table, store = store_path)
lyte_tbl_war <- pick_case_table(lyte_tbl_all, case_name)

lyte_global_all <- tar_read(lyte_global_tests, store = store_path)
lyte_global_war <- pick_case_table(lyte_global_all, case_name)


# figures
km_vec <- tar_read(km_plot_file, store = store_path)
km_file <- km_vec[grepl("/war/", km_vec, fixed = TRUE)][1]
km_file <- normalizePath(km_file, mustWork = TRUE)

illeg_vec <- tar_read(illegitimacy_plot_file, store = store_path)
illeg_file <- illeg_vec[grepl("/war/", illeg_vec, fixed = TRUE)][1]
illeg_file <- normalizePath(illeg_file, mustWork = TRUE)

lyte_vec <- tar_read(lyte_prediction_plot_file, store = store_path)
lyte_file <- lyte_vec[grepl("/war/", lyte_vec, fixed = TRUE)][1]
lyte_file <- normalizePath(lyte_file, mustWork = TRUE)

sex_ratio_file <- tar_read(sex_ratio_plot_file, store = store_path)
sex_ratio_file <- normalizePath(sex_ratio_file, mustWork = TRUE)

panel_file <- tar_read(sex_stillbirth_panel_plot_file, store = "_targets")
panel_file <- normalizePath(panel_file, mustWork = TRUE)

boy_share_live_still_file <- tar_read(boy_share_live_still_plot_file, store = "_targets")
boy_share_live_still_file <- normalizePath(boy_share_live_still_file, mustWork = TRUE)

stopifnot(file.exists(km_file))
stopifnot(file.exists(illeg_file))
stopifnot(file.exists(lyte_file))

# formatted tables
lyte_tbl_war_fmt <- format_lyte_table(lyte_tbl_war, digits = 3)
if ("term" %in% names(lyte_tbl_war_fmt)) {
  lyte_tbl_war_fmt[, term := clean_term_labels(term)]
}

lyte_global_war_fmt <- format_global_tests(lyte_global_war, digits = 2)
if ("term" %in% names(lyte_global_war_fmt)) {
  lyte_global_war_fmt[, term := clean_term_labels(term)]
}
```

# Introduction

Research on the demographic consequences of war has traditionally concentrated on excess
mortality, particularly among soldiers and civilian populations directly exposed to violence, famine,
and epidemic disease. Less attention has been paid to how armed conflict affects reproductive
behavior and early-life conditions for children conceived and born during periods of crisis. This
paper shifts the focus from mortality to fertility and early-life outcomes, examining how the Finnish
War (1808–1809) affected women and children in northern Sweden, with particular attention to
Västerbotten and the Umeå region.

The Finnish War, fought between Sweden and Russia, culminated in the loss of Finland to the
Russian Empire and temporarily transformed parts of northern Sweden into a military frontier.
Umeå and neighboring parishes became sites of troop movements, temporary encampments, and
improvised military hospitals as Swedish forces retreated westward. Civilian communities were
subjected to food requisitions, local disruption, and the spread of infectious disease. While the
immediate mortality crisis has been documented, the longer-term demographic consequences for
reproductive behavior and child outcomes remain underexplored.

Using longitudinal parish register data, this study investigates three interrelated questions:

First, did fertility decline during the war? Periods of military occupation, forced requisitions, and
male inscription in the army may have disrupted marriage patterns and reduced conception rates. At
the same time, stress, malnutrition, and disease exposure could have affected fertility and pregnancy
outcomes.

Second, did the composition of births change? We examine whether the proportion of non-marital
births increased during the conflict, potentially reflecting disrupted social control, delayed marriage,
altered partnership patterns, or asymmetric power relations between soldiers and civilian women.
Spatial variation across parishes with differing levels of military presence allows us to evaluate
whether such changes were locally concentrated.

Third, what were the long-term consequences for children born during or immediately after the
war? Using a life-course perspective, we follow exposed cohorts to assess infant and child
mortality, later-life survival, and indicators of disability and socioeconomic disadvantage. Here we
use the Developmental Origins of Health and Disease (DOHaD) framework, which emphasizes how
conditions in utero and early childhood can have lasting effects on health, survival, and
socioeconomic outcomes.
We expect to observe a marked decline in fertility during the most intense phases of the conflict,
followed by a post-war rebound indicative of postponed rather than permanently foregone
childbearing. We further anticipate an increase in the proportion of non-marital births in parishes
characterized by sustained military presence, reflecting disrupted marriage patterns and weakened
local social regulation. Finally, we hypothesize that children conceived or born during the war will
display elevated early-life mortality and, among survivors, signs of longer-term scarring consistent
with early-life exposure to nutritional stress and infectious disease.
By centering women and children rather than soldiers, this paper contributes to a more
comprehensive understanding of war as a demographic shock. The Finnish War serves as a natural
experiment illustrating how armed conflict reshapes fertility behavior, and life chances for cohorts
born under crisis. These findings underscore the importance of incorporating reproductive and
early-life perspectives into historical studies of war and population health.

#  Historical context

The Finnish War disrupted economic and social life across northern Sweden. Military mobilisation, institutional breakdown, and potential food shortages may have affected early-life conditions.



#  Data

The data are derived from the POPUM database and consist of linked individual-level records including birth, death, residence, and health indicators.

Individuals are constructed from multiple residence spells, with adjustments for overlaps and gaps. Exit types are used to distinguish mortality from migration and censoring.

##  Descriptive statistics

```{r}
kable(desc_war)
```

# Methods

We use two complementary approaches:

- Kaplan–Meier survival analysis
- Logistic regression of disability before age 20

Cohorts are defined within a narrow time window around the Finnish War.

#  Results

##  Survival

```{r, out.width="75%", fig.align="center"}
# knitr::include_graphics(km_file)
image_md(km_file)
```



##  Legitimacy patterns
Crisis changes who gets a child, more unmarried?, more farmers? Results in selection bias. 

```{r, out.width="75%", fig.align="center"}
# knitr::include_graphics(illeg_file)
image_md(illeg_file)
```

## Sex ratio
Crisis might result in selection in uterus. Boys might be more vulnerable to crisis.

```{r, out.width="75%", fig.align="center"}
image_md(sex_ratio_file)
```


## Still births

```{r, out.width="80%", fig.align="center"}
image_md(panel_file)
```

```{r, out.width="80%", fig.align="center"}
image_md(boy_share_live_still_file)
```

 

## Health outcomes, risk of disability

```{r, out.width="75%", fig.align="center"}
# knitr::include_graphics(lyte_file)
# image_md(lyte_file)
```

###  Model estimates

```{r}
kable(
  lyte_tbl_war_fmt[
    ,
    .(term, estimate, std.error, statistic, p.value, odds_ratio, conf.low.or, conf.high.or)
  ],
  col.names = c("Term", "Estimate", "SE", "z", "p", "Odds ratio", "CI low", "CI high")
)
```

##  Global tests for model terms

```{r}
kable(
  lyte_global_war_fmt[
    ,
    .(term, df, statistic, p.value)
  ],
  col.names = c("Term", "Df", "LR Chi-square", "p")
)
```

#  Discussion

The results suggest that cohorts born during the Finnish War experienced altered early-life conditions with measurable consequences for both survival and health.

The observed patterns are consistent with a combination of selection effects and direct causal impacts.

#  Conclusion

This study demonstrates that short-term crises can have lasting demographic effects. By combining detailed longitudinal data with cohort-based identification, we provide new evidence on the importance of early-life environments.

## Acknowledgements {-}

Optional.

## Notes {-}

Optional.
