---
title: "Born under crisis: Early-life conditions and health outcomes during the Finnish War"
author:
  - name: "Erling Häggström Gunfridsson"
    affiliation: "Umeå University"
  - name: "Elina Nordun Omnell"
    affiliation: "Umeå University"
date: today
format:
  html:
    toc: true
    number-sections: true
  pdf:
    toc: true
    number-sections: true
    documentclass: article
    geometry: margin=2.5cm
    fontsize: 11pt
execute:
  echo: false
  warning: false
  message: false
---

## Abstract {-}

This paper examines the long-term consequences of early-life exposure to crisis conditions during the Finnish War (1808–1809). Using longitudinal individual-level data from northern Sweden, we analyse both survival and early-life health outcomes. The results indicate that cohorts born during the crisis exhibit distinct patterns in mortality and disability, consistent with both selection effects and causal impacts of adverse early-life environments.

**Keywords:** early-life conditions, mortality, health, Finnish War, historical demography

```{r}
library(targets)
library(data.table)
library(ggplot2)
library(knitr)

pick_case_table <- function(x, case_name = "war") {
  dt <- data.table::as.data.table(x)
  if (!"case" %in% names(dt)) return(dt)
  out <- dt[case == case_name]
  if (nrow(out) == 0L) {
    stop(sprintf("No rows found for case '%s'.", case_name), call. = FALSE)
  }
  out
}

pick_case_file <- function(x, case_name = "war") {
  if (is.character(x)) {
    hit <- grepl(paste0("/", case_name, "/"), x, fixed = TRUE)
    if (any(hit)) return(unname(x[hit][1]))
    nms <- names(x)
    if (!is.null(nms)) {
      hit_nms <- grepl(case_name, nms, fixed = TRUE)
      if (any(hit_nms)) return(unname(x[hit_nms][1]))
    }
    if (length(x) == 1L) return(unname(x[1]))
  }

  dt <- data.table::as.data.table(x)
  if ("case" %in% names(dt)) {
    out <- dt[case == case_name]
    if (nrow(out) == 0L) {
      stop(sprintf("No file found for case '%s'.", case_name), call. = FALSE)
    }
  } else {
    out <- dt
  }

  for (nm in c("path", "file", "value")) {
    if (nm %in% names(out)) return(out[[nm]][[1]])
  }

  stop("Could not identify the correct file for the selected case.", call. = FALSE)
}

format_p_value <- function(p) {
  ifelse(
    is.na(p),
    NA_character_,
    ifelse(p < 0.001, "<0.001", sprintf("%.3f", p))
  )
}

format_lyte_table <- function(dt, digits = 3) {
  dt <- data.table::copy(data.table::as.data.table(dt))

  keep <- intersect(
    c(
      "case", "term", "estimate", "std.error", "statistic", "p.value",
      "odds_ratio", "conf.low.or", "conf.high.or"
    ),
    names(dt)
  )
  dt <- dt[, ..keep]

  if ("term" %in% names(dt)) {
    dt <- dt[term != "(Intercept)"]
  }

  num_cols <- intersect(
    c("estimate", "std.error", "statistic", "odds_ratio", "conf.low.or", "conf.high.or"),
    names(dt)
  )
  if (length(num_cols) > 0) {
    dt[, (num_cols) := lapply(.SD, round, digits), .SDcols = num_cols]
  }

  if ("p.value" %in% names(dt)) {
    dt[, p.value := format_p_value(p.value)]
  }

  dt
}

format_global_tests <- function(dt, digits = 2) {
  dt <- data.table::copy(data.table::as.data.table(dt))

  keep <- intersect(c("case", "term", "df", "statistic", "p.value"), names(dt))
  dt <- dt[, ..keep]

  num_cols <- intersect(c("df", "statistic"), names(dt))
  if (length(num_cols) > 0) {
    dt[, (num_cols) := lapply(.SD, round, digits), .SDcols = num_cols]
  }

  if ("p.value" %in% names(dt)) {
    dt[, p.value := format_p_value(p.value)]
  }

  dt
}

clean_term_labels <- function(x) {
  x <- gsub("^factor\\(fod_ar\\)", "Birth year: ", x)
  x <- gsub("^fod_ar", "Birth year", x)
  x <- gsub("^oaktaTRUE$", "Illegitimate birth", x)
  x <- gsub("^oakta$", "Illegitimate birth", x)
  x <- gsub("^kon$", "Sex", x)
  x <- gsub("^kon1$", "Sex: category 1", x)
  x <- gsub("^kon2$", "Sex: category 2", x)
  x
}

case_name <- "war"
store_path <- "../_targets"

desc_all <- tar_read(descriptive_table, store = store_path)
desc_war <- pick_case_table(desc_all, case_name)

lyte_tbl_all <- tar_read(lyte_model_table, store = store_path)
lyte_tbl_war <- pick_case_table(lyte_tbl_all, case_name)

lyte_global_all <- tar_read(lyte_global_tests, store = store_path)
lyte_global_war <- pick_case_table(lyte_global_all, case_name)

km_file <- pick_case_file(tar_read(km_plot_file, store = store_path), case_name)
km_file <- file.path("..", km_file)

illeg_file <- pick_case_file(tar_read(illegitimacy_plot_file, store = store_path), case_name)
illeg_file <- file.path("..", illeg_file)

lyte_file <- pick_case_file(tar_read(lyte_prediction_plot_file, store = store_path), case_name)
lyte_file <- file.path("..", lyte_file)

stopifnot(file.exists(km_file))
stopifnot(file.exists(illeg_file))
stopifnot(file.exists(lyte_file))

lyte_tbl_war_fmt <- format_lyte_table(lyte_tbl_war, digits = 3)
if ("term" %in% names(lyte_tbl_war_fmt)) {
  lyte_tbl_war_fmt[, term := clean_term_labels(term)]
}

lyte_global_war_fmt <- format_global_tests(lyte_global_war, digits = 2)
if ("term" %in% names(lyte_global_war_fmt)) {
  lyte_global_war_fmt[, term := clean_term_labels(term)]
}
```

# 1. Introduction

Early-life conditions are widely recognised as key determinants of long-term health and survival. Periods of crisis, such as war or famine, provide exogenous variation in living conditions and therefore offer an opportunity to study these effects.

This paper examines cohorts born around the Finnish War (1808–1809) in northern Sweden. We analyse both survival and early-life health outcomes, measured through disability before age 20.

# 2. Historical context

The Finnish War disrupted economic and social life across northern Sweden. Military mobilisation, institutional breakdown, and potential food shortages may have affected early-life conditions.

Peripheral regions such as Västerbotten are particularly suitable for studying these effects due to their relative isolation and dependence on local production.

# 3. Data

The data are derived from the POPUM database and consist of linked individual-level records including birth, death, residence, and health indicators.

Individuals are constructed from multiple residence spells, with adjustments for overlaps and gaps. Exit types are used to distinguish mortality from migration and censoring.

## 3.1 Descriptive statistics

```{r}
kable(desc_war)
```

# 4. Methods

We use two complementary approaches:

- Kaplan–Meier survival analysis
- Logistic regression of disability before age 20

Cohorts are defined within a narrow time window around the Finnish War.

# 5. Results

## 5.1 Survival

```{r, out.width="75%", fig.align="center"}
knitr::include_graphics(km_file)
```

## 5.2 Legitimacy patterns

```{r, out.width="75%", fig.align="center"}
knitr::include_graphics(illeg_file)
```

## 5.3 Health outcomes

```{r, out.width="75%", fig.align="center"}
knitr::include_graphics(lyte_file)
```

## 5.4 Model estimates

```{r}
kable(
  lyte_tbl_war_fmt[
    ,
    .(term, estimate, std.error, statistic, p.value, odds_ratio, conf.low.or, conf.high.or)
  ],
  col.names = c("Term", "Estimate", "SE", "z", "p", "Odds ratio", "CI low", "CI high")
)
```

## 5.5 Global tests for model terms

```{r}
kable(
  lyte_global_war_fmt[
    ,
    .(term, df, statistic, p.value)
  ],
  col.names = c("Term", "Df", "LR Chi-square", "p")
)
```

# 6. Discussion

The results suggest that cohorts born during the Finnish War experienced altered early-life conditions with measurable consequences for both survival and health.

The observed patterns are consistent with a combination of selection effects and direct causal impacts.

# 7. Conclusion

This study demonstrates that short-term crises can have lasting demographic effects. By combining detailed longitudinal data with cohort-based identification, we provide new evidence on the importance of early-life environments.

## Acknowledgements {-}

Optional.

## Notes {-}

Optional.
